---
title: "SBC Bioreactors Conditioning"
author: "Nicholas Baetge"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(ggrepel)

data_path <- "~/github/sbc_bioreactors/data/bioreactor_data.xlsx"
detection_limit <- 1.5  # µmol C L-1

water_map <- c(
  "UCSBintake"  = "UCSB intake",
  "goletapier"  = "Goleta Pier",
  "stearnswarf" = "Stearns Wharf",
  "sbc"         = "Santa Barbara Channel"
)

trt_map <- c(
  "flowthru"      = "ambient",
  "DiatomExudate" = "diatom exudate",
  "DiatomLysate"  = "diatom lysate",
  "0m"            = "surface",
  "0 m"           = "surface",
  "75m"           = "75 m",
  "200m"          = "200 m"
)

df <- read_excel(data_path, sheet = 1) %>%
  mutate(
    date = as.Date(as.character(ymd), "%Y%m%d"),
    water_lab = recode(water, !!!water_map, .default = water),
    trt_lab = recode(trt, !!!trt_map, .default = trt),
    stage = as.integer(bioreactor_n),

    # Raw flowrate values are stored in mL s-1.
    flow_rate_ml_min = flowrate * 60,
    flow_rate_ml_min = case_when(
      is.na(flow_rate_ml_min) ~ NA_real_,
      abs(flow_rate_ml_min - 4) < 0.1 ~ 4,
      abs(flow_rate_ml_min - 20) < 0.1 ~ 20,
      TRUE ~ flow_rate_ml_min
    ),

    # Recover analytical vial replication only when
    # SE = SD / sqrt(n) yields an integer n.
    n_doc_vials_raw = if_else(
      is.finite(se_doc) & se_doc > 0,
      (sd_doc / se_doc)^2,
      NA_real_
    ),

    n_doc_vials = case_when(
      # In this workbook, missing or zero SE denotes a single DOC sample.
      # Some of these rows retain a nonzero SD value from upstream processing,
      # so SE == 0 is treated as the decisive single-sample flag.
      is.na(se_doc) | se_doc == 0 ~ 1L,

      # Otherwise recover analytical vial replication when
      # SE = SD / sqrt(n) yields an integer n.
      is.finite(n_doc_vials_raw) &
        abs(n_doc_vials_raw - round(n_doc_vials_raw)) < 1e-5 ~
        as.integer(round(n_doc_vials_raw)),

      TRUE ~ NA_integer_
    )
  ) %>%
  filter(is.finite(mean_doc))

write_csv(
  df,
  "results/df.csv"
)

# Missing or zero SE is accepted as the workbook's single-vial flag.
# Stop only when SE is positive but SD/SE does not support a valid integer n.
n_bad <- df %>%
  filter(
    is.finite(se_doc),
    se_doc > 0,
    is.na(n_doc_vials)
  ) %>%
  nrow()

if (n_bad > 0) {
  stop(
    n_bad,
    " DOC rows with positive SE contain SD/SE information that does not yield an integer vial count."
  )
}

doc_vial_check <- df %>%
  count(n_doc_vials, name = "n_rows") %>%
  arrange(n_doc_vials)

print(doc_vial_check)


single_vial_rows <- df %>%
  filter(n_doc_vials == 1) %>%
  select(
    exp, date, water_lab, trt_lab, type,
    bioreactor, stage, hours_elapsed,
    mean_doc, sd_doc, se_doc, n_doc_vials
  )

print(single_vial_rows)

write_csv(
  single_vial_rows,
  "results/single_vial_DOC_rows.csv"
)

dir.create("results", showWarnings = FALSE)
dir.create("figures", showWarnings = FALSE)
```

# Replication and metadata audit

```{r replication-audit}
replication_audit <- df %>%
  count(
    exp, date, water_lab, trt_lab, stage, bioreactor,
    flow_rate_ml_min, hours_elapsed,
    name = "duplicate_key_count"
  ) %>%
  arrange(desc(duplicate_key_count), exp, date)

# With flow rate included, BR15-4 consists of two distinct assay conditions:
# 4 mL min-1 and 20 mL min-1.
replication_audit %>%
  filter(duplicate_key_count > 1)

doc_vial_summary <- df %>%
  count(n_doc_vials, name = "n_rows") %>%
  arrange(n_doc_vials)

write_csv(replication_audit, "results/replication_audit.csv")
write_csv(doc_vial_summary, "results/doc_vial_replication_summary.csv")
```

# Figure 2: conditioning trajectory

The conditioning figure uses the same four original columns (1, 3, 5, and 7), ambient UCSB intake water, and first-stage effluent. Individual column removals are shown directly. Experiment-level points and error bars represent the arithmetic mean ± SD among columns.

The experiments are displayed as discrete conditioning assays rather than as a continuous line series. BR15-1 is included as the final conditioning assay, approximately 5.65 months (172 days) after BR14-1. Because no recurring labile-carbon positive control was used, the figure is interpreted operationally as the interval before reproducible DOC removal was observed, not as proof that biofilm maturation alone caused the transition.

For each experiment, the influent bar shows the measured 0 h DOC concentration and its analytical SD. The effluent bar shows the arithmetic mean ± SD among the four independent first-stage bioreactor columns. Individual effluent-column measurements are overlaid as points. Positive DOC-removal labels indicate that effluent DOC was lower than influent DOC.



```{r figure-2-data}
conditioning_experiments <- c(paste0("BR14-", 1:8), "BR15-1")

cond <- df %>%
  filter(
    exp %in% conditioning_experiments,
    type == "bioreactor",
    water == "UCSBintake",
    trt == "flowthru",
    hours_elapsed %in% c(0, 4),
    stage %in% c(0, 1)
  )

cond_in <- cond %>%
  filter(stage == 0, hours_elapsed == 0) %>%
  select(
    exp, date,
    influent_mean = mean_doc,
    influent_sd = sd_doc,
    influent_se = se_doc,
    influent_n_vials = n_doc_vials
  ) %>%
  distinct()

cond_eff <- cond %>%
  filter(
    stage == 1,
    hours_elapsed == 4,
    bioreactor %in% c(1, 3, 5, 7)
  ) %>%
  select(
    exp, date, bioreactor, beads,
    flow_rate_ml_min,
    effluent_mean = mean_doc,
    effluent_sd = sd_doc,
    effluent_se = se_doc,
    effluent_n_vials = n_doc_vials
  )

conditioning_reactor <- cond_eff %>%
  left_join(cond_in, by = c("exp", "date")) %>%
  mutate(
    conditioning_start = min(date),
    days_conditioning = as.integer(date - conditioning_start),
    months_conditioning = days_conditioning / 30.4375,
    doc_removal = influent_mean - effluent_mean,
    propagated_se_delta = sqrt(influent_se^2 + effluent_se^2),
    z_value = if_else(
      is.finite(propagated_se_delta) & propagated_se_delta > 0,
      doc_removal / propagated_se_delta,
      NA_real_
    ),
    p_value = if_else(
      is.finite(z_value),
      2 * pnorm(-abs(z_value)),
      NA_real_
    ),
    exceeds_detection_limit = doc_removal >= detection_limit
  )

conditioning_summary <- conditioning_reactor %>%
  group_by(exp, date, days_conditioning, months_conditioning) %>%
  summarise(
    influent_mean = first(influent_mean),
    influent_sd = first(influent_sd),
    influent_n_vials = first(influent_n_vials),
    mean_effluent_doc = mean(effluent_mean),
    sd_effluent_doc = if_else(n() > 1, sd(effluent_mean), NA_real_),
    mean_doc_removal = influent_mean - mean_effluent_doc,
    sd_doc_removal = sd(doc_removal),
    n_bioreactors = n(),
    n_exceeding_detection_limit = sum(exceeds_detection_limit),
    effluent_n_vials_min = min(effluent_n_vials, na.rm = TRUE),
    effluent_n_vials_max = max(effluent_n_vials, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(date) %>%
  mutate(
    exp = factor(exp, levels = conditioning_experiments),
    x_label = paste0(
      as.character(exp), "\n",
      format(date, "%Y-%m-%d"), "\n",
      sprintf("%.2f months", months_conditioning)
    ),
    x_label = factor(x_label, levels = x_label),
    removal_label = paste0(
      sprintf("%.1f", mean_doc_removal),
      "\n",
      "n=", n_bioreactors
    ) 
  ) 

conditioning_summary <- conditioning_summary %>%
  arrange(date) %>%
  mutate(
    exp_label = factor(
      as.character(exp),
      levels = conditioning_experiments
    ),
    month_label = sprintf("%.1f mo", months_conditioning),
    axis_label = paste0(as.character(exp_label), "\n", month_label)
  )

conditioning_plot_data <- conditioning_summary %>%
  select(
    exp, date, exp_label, months_conditioning,
    influent_mean, influent_sd,
    mean_effluent_doc, sd_effluent_doc
  ) %>%
  pivot_longer(
    cols = c(influent_mean, mean_effluent_doc),
    names_to = "stream",
    values_to = "mean_doc"
  ) %>%
  mutate(
    stream = recode(
      stream,
      "influent_mean" = "Influent (0 h)",
      "mean_effluent_doc" = "Effluent (4 h)"
    ),
    stream = factor(
      stream,
      levels = c("Influent (0 h)", "Effluent (4 h)")
    ),
    sd_doc_plot = if_else(
      stream == "Influent (0 h)",
      influent_sd,
      sd_effluent_doc
    )
  )

conditioning_points <- conditioning_reactor %>%
  left_join(
    conditioning_summary %>%
      select(exp, exp_label),
    by = "exp"
  ) %>%
  mutate(
    stream = factor(
      "Effluent (4 h)",
      levels = c("Influent (0 h)", "Effluent (4 h)")
    )
  )

write_csv(
  conditioning_reactor,
  "results/figure2_conditioning_reactor_level.csv"
)

write_csv(
  conditioning_summary,
  "results/figure2_conditioning_summary.csv"
)
```

The experiments are displayed as discrete conditioning assays rather than as a continuous line series. BR15-1 is included as the final conditioning assay, approximately 5.65 months (172 days) after BR14-1.

For each experiment, the influent bar shows the measured 0 h DOC concentration and its analytical SD. The effluent bar shows the arithmetic mean ± SD among the four independent first-stage bioreactor columns. Individual effluent-column measurements are overlaid as points. Positive DOC-removal labels indicate that effluent DOC was lower than influent DOC.

```{r figure-2, fig.width=8.2, fig.height=5.2}
pd2 <- position_dodge(width = 0.55)

fig2_ymax <- max(
  conditioning_plot_data$mean_doc +
    coalesce(conditioning_plot_data$sd_doc_plot, 0),
  na.rm = TRUE
)

fig2_ymin <- min(
  conditioning_plot_data$mean_doc -
    coalesce(conditioning_plot_data$sd_doc_plot, 0),
  na.rm = TRUE
)

fig2_label_offset <- 0.05 * (fig2_ymax - fig2_ymin)

axis_labels <- conditioning_summary %>%
  select(exp_label, axis_label) %>%
  distinct() %>%
  deframe()

fig2 <- ggplot(
  conditioning_plot_data,
  aes(
    x = exp_label,
    y = mean_doc,
    group = stream
  )
) +

  # Mean ± SD for influent and effluent
  geom_errorbar(
    aes(
      ymin = mean_doc - sd_doc_plot,
      ymax = mean_doc + sd_doc_plot
    ),
    position = pd2,
    width = 0.12,
    linewidth = 0.65,
    na.rm = TRUE,
    show.legend = FALSE
  ) +

  # Summary means: fill denotes stream; fixed square shape
  geom_point(
    aes(fill = stream),
    shape = 22,
    position = pd2,
    size = 3.7,
    stroke = 0.7,
    color = "black",
    show.legend = TRUE
  ) +

  # Individual effluent columns: shape denotes bead packing
  geom_point(
    data = conditioning_points,
    aes(
      x = exp_label,
      y = effluent_mean,
      shape = beads
    ),
    inherit.aes = FALSE,
    position = position_nudge(x = 0.1375),
    size = 2.3,
    alpha = 0.75,
    color = "grey30",
    show.legend = TRUE
  ) +

  # Separate BR15-1 visually
  geom_vline(
    xintercept = 8.5,
    linetype = "dashed",
    linewidth = 0.4
  ) +

  annotate(
    "text",
    x = 9,
    y = fig2_ymax + fig2_label_offset,
    label = "First clear removal",
    size = 3.3,
    fontface = "italic"
  ) +

  scale_x_discrete(
    labels = axis_labels
  ) +

  # Only stream appears in the fill legend
  scale_fill_manual(
    values = c(
      "Influent (0 h)" = "grey75",
      "Effluent (4 h)" = "#56B4E9"
    ),
    labels = c(
      "Influent (0 h)" = "Influent (0 h)",
      "Effluent (4 h)" = "Effluent (4 h)"
    ),
    name = NULL
  ) +

  # Only bead configuration appears in the shape legend
  scale_shape_manual(
    values = c(
      "homogenous" = 16,
      "mixed" = 17
    ),
    labels = c(
      "homogenous" = "Homogeneous (1 mm)",
      "mixed" = "Mixed (1 + 4 mm)"
    ),
    name = NULL
  ) +

  guides(
    fill = guide_legend(
      order = 1,
      title = NULL,
      override.aes = list(
        shape = 22,
        size = 4,
        color = "black",
        alpha = 1
      )
    ),
    shape = guide_legend(
      order = 2,
      title = NULL,
      override.aes = list(
        size = 2.8,
        color = "grey30",
        alpha = 1
      )
    )
  ) +

  labs(
    x = NULL,
    y = expression("DOC (" * mu * "mol C L"^{-1} * ")"),
    # caption = paste(
    #   "Large squares show mean DOC \u00B1 SD.",
    #   "Small symbols show individual effluent bioreactor columns.",
    #   "Influent SD reflects analytical replication;",
    #   "effluent SD reflects variation among columns."
    # )
  ) +

  coord_cartesian(
    ylim = c(
      floor(fig2_ymin - 0.5),
      ceiling(fig2_ymax + 2 * fig2_label_offset)
    ),
    clip = "off"
  ) +

  theme_bw(base_size = 11) +

  theme(
    legend.position = "top",
    legend.box = "vertical",
    legend.margin = margin(0, 0, 2, 0),
    legend.box.margin = margin(0, 0, 0, 0),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1,
      lineheight = 0.95
    ),
    plot.caption = element_text(
      hjust = 0,
      size = 9
    ),
    plot.margin = margin(6, 12, 12, 6)
  )

fig2

ggsave(
  "figures/Figure_2_conditioning.pdf",
  fig2,
  width = 11,
  height = 6.5
)
```

# Figure 3: post-conditioning single-column performance

The experimental unit is an individual first-stage bioreactor column. Each plotted point is the DOC removal measured for one column. Group means and error bars represent the arithmetic mean ± SD among columns.

The number of DOC vials underlying each concentration estimate is retained separately and is not combined with the number of bioreactor columns.

For BR15-4, flow rate is treated as an explicit experimental factor. The 4 and 20 mL min-1 observations are therefore analyzed and displayed as separate conditions.

Four- and seven-hour sampling times represent the time required to flush the corresponding configuration several residence times, rather than different biological exposure durations.

```{r figure-3-data}
# Figure 3: post-conditioning single-column performance

post <- df %>%
  filter(
    !exp %in% c(paste0("BR14-", 1:8), "BR15-1"),
    type == "bioreactor",
    stage %in% c(0, 1),
    hours_elapsed %in% c(0, 4, 7)
  ) %>%
  mutate(
    beads = tolower(trimws(beads)),
    beads = recode(
      beads,
      "homogenous" = "homogeneous",
      "mixed 1+4mm" = "mixed",
      .default = beads
    ),
    beads = factor(
      beads,
      levels = c("homogeneous", "mixed")
    )
  )

# Influent rows do not have a reactor-specific flow rate.
# They are joined to each corresponding effluent condition by
# experiment, date, source water, and treatment.
post_in <- post %>%
  filter(
    stage == 0,
    hours_elapsed == 0
  ) %>%
  select(
    exp, date, water_lab, trt_lab,
    influent_mean = mean_doc,
    influent_sd = sd_doc,
    influent_se = se_doc,
    influent_n_vials = n_doc_vials
  ) %>%
  distinct()

post_eff <- post %>%
  filter(
    stage == 1,
    hours_elapsed %in% c(4, 7)
  ) %>%
  select(
    exp, date, water_lab, trt_lab,
    flow_rate_ml_min, hours_elapsed,
    bioreactor, beads,
    effluent_mean = mean_doc,
    effluent_sd = sd_doc,
    effluent_se = se_doc,
    effluent_n_vials = n_doc_vials
  )

post_reactor <- post_eff %>%
  left_join(
    post_in,
    by = c("exp", "date", "water_lab", "trt_lab")
  ) %>%
  mutate(
    doc_removal = influent_mean - effluent_mean,
    pct_removal = 100 * doc_removal / influent_mean,

    source_treatment = case_when(
      trt_lab %in% c("75 m", "200 m") ~ paste(water_lab, trt_lab),
      trt_lab == "surface" ~ paste(water_lab, "surface"),
      trt_lab == "ambient" ~ water_lab,
      TRUE ~ paste(water_lab, trt_lab, sep = " + ")
    ),

    flow_label = case_when(
      is.na(flow_rate_ml_min) ~ "flow not recorded",
      TRUE ~ paste0(format(flow_rate_ml_min, trim = TRUE), " mL min\u207B\u00B9")
    ),

    # Only append flow rate to the displayed experiment label when
    # more than one flow condition occurred within that experiment.
    experiment_condition = exp
  )

multi_flow_experiments <- post_reactor %>%
  filter(!is.na(flow_rate_ml_min)) %>%
  distinct(exp, flow_rate_ml_min) %>%
  count(exp, name = "n_flow_rates") %>%
  filter(n_flow_rates > 1) %>%
  pull(exp)

post_reactor <- post_reactor %>%
  mutate(
    experiment_condition = if_else(
      exp %in% multi_flow_experiments,
      paste0(exp, "\n", flow_label),
      exp
    )
  )

post_summary <- post_reactor %>%
  group_by(
    exp, experiment_condition, date,
    source_treatment, flow_rate_ml_min, flow_label,
    hours_elapsed
  ) %>%
  summarise(
    mean_doc_removal = mean(doc_removal, na.rm = TRUE),

    sd_doc_removal = if_else(
      sum(is.finite(doc_removal)) > 1,
      sd(doc_removal, na.rm = TRUE),
      NA_real_
    ),

    mean_pct_removal = mean(pct_removal, na.rm = TRUE),

    sd_pct_removal = if_else(
      sum(is.finite(pct_removal)) > 1,
      sd(pct_removal, na.rm = TRUE),
      NA_real_
    ),

    n_bioreactors = n_distinct(
      interaction(bioreactor, flow_rate_ml_min, drop = TRUE)
    ),

    n_unique_bioreactor_labels = n_distinct(bioreactor),

    influent_n_vials = first(influent_n_vials),

    effluent_n_vials_min = if_else(
      any(is.finite(effluent_n_vials)),
      min(effluent_n_vials, na.rm = TRUE),
      NA_integer_
    ),

    effluent_n_vials_max = if_else(
      any(is.finite(effluent_n_vials)),
      max(effluent_n_vials, na.rm = TRUE),
      NA_integer_
    ),

    .groups = "drop"
  )

# Arrange experiments in the same analytical order used in Table 1.
experiment_order <- c(
  "BR15-3",
  "BR15-5",
  "BR15-6",
  "BR15-7",
  "BR15-4\n4 mL min\u207B\u00B9",
  "BR15-4\n20 mL min\u207B\u00B9",
  "BR15-10",
  "BR17-1",
  "BR16-2",
  "BR17-2",
  "BR15-11",
  "BR16-1",
  "BR16-3"
)

post_reactor <- post_reactor %>%
  mutate(
    experiment_condition = factor(
      experiment_condition,
      levels = experiment_order
    )
  )

post_reactor <- post_reactor %>%
  mutate(
    site = case_when(
      water_lab == "UCSB intake" ~ "UCSB intake",
      water_lab == "Goleta Pier" ~ "Goleta Pier",
      water_lab == "Stearns Wharf" ~ "Stearns Wharf",
      water_lab == "Santa Barbara Channel" ~ "Santa Barbara Channel",
      water_lab == "SBC" ~ "Santa Barbara Channel",
      TRUE ~ water_lab
    ),

    condition = case_when(
      trt_lab == "diatom lysate" ~ "Diatom lysate",
      trt_lab == "75 m" ~ "75 m",
      trt_lab == "200 m" ~ "200 m",
      TRUE ~ "Surface/ambient"
    ),

    site = factor(
      site,
      levels = c(
        "UCSB intake",
        "Goleta Pier",
        "Stearns Wharf",
        "Santa Barbara Channel"
      )
    ),

    condition = factor(
      condition,
      levels = c(
        "Surface/ambient",
        "Diatom lysate",
        "75 m",
        "200 m"
      )
    )
  )

post_summary <- post_reactor %>%
  group_by(
    exp,
    experiment_condition,
    date,
    site,
    condition,
    flow_rate_ml_min,
    flow_label,
    hours_elapsed
  ) %>%
  summarise(
    mean_doc_removal = mean(doc_removal, na.rm = TRUE),

    sd_doc_removal = if_else(
      sum(is.finite(doc_removal)) > 1,
      sd(doc_removal, na.rm = TRUE),
      NA_real_
    ),

    mean_pct_removal = mean(pct_removal, na.rm = TRUE),

    sd_pct_removal = if_else(
      sum(is.finite(pct_removal)) > 1,
      sd(pct_removal, na.rm = TRUE),
      NA_real_
    ),

    n_bioreactors = n(),

    n_unique_bioreactor_labels =
      n_distinct(bioreactor),

    influent_n_vials =
      first(influent_n_vials),

    effluent_n_vials_min = if_else(
      any(is.finite(effluent_n_vials)),
      min(effluent_n_vials, na.rm = TRUE),
      NA_integer_
    ),

    effluent_n_vials_max = if_else(
      any(is.finite(effluent_n_vials)),
      max(effluent_n_vials, na.rm = TRUE),
      NA_integer_
    ),

    .groups = "drop"
  )

write_csv(
  post_reactor,
  "results/figure3_postconditioning_reactor_level.csv"
)

write_csv(
  post_summary,
  "results/figure3_postconditioning_summary.csv"
)
```

```{r figure-3, fig.width=11, fig.height=6.5}

pd3 <- position_dodge(width = 0.55)

fig3 <- ggplot(
  post_reactor,
  aes(
    x = experiment_condition,
    y = doc_removal,
    group = condition
  )
) +

  geom_hline(
    yintercept = 0,
    linetype = "dashed",
    linewidth = 0.45
  ) +

  geom_hline(
    yintercept = detection_limit,
    linetype = "dotted",
    linewidth = 0.45
  ) +

  # Mean ± SD among columns
  geom_errorbar(
    data = post_summary,
    aes(
      x = experiment_condition,
      ymin = mean_doc_removal - sd_doc_removal,
      ymax = mean_doc_removal + sd_doc_removal,
      group = condition
    ),
    inherit.aes = FALSE,
    position = pd3,
    width = 0.12,
    linewidth = 0.55,
    na.rm = TRUE,
    show.legend = FALSE
  ) +

  # Individual bioreactor columns
  geom_point(
    aes(
      shape = beads,
      fill = condition
    ),
    position = position_jitterdodge(
      jitter.width = 0.06,
      jitter.height = 0,
      dodge.width = 0.55
    ),
    size = 2.5,
    color = "grey25",
    stroke = 0.5,
    alpha = 0.8,
    show.legend = TRUE
  ) +

  # Group means
  geom_point(
    data = post_summary,
    aes(
      x = experiment_condition,
      y = mean_doc_removal,
      fill = condition,
      group = condition
    ),
    inherit.aes = FALSE,
    position = pd3,
    shape = 22,
    color = "black",
    size = 3.6,
    stroke = 0.8,
    show.legend = FALSE
  ) +

  facet_wrap(
    ~ site,
    ncol = 2,
    scales = "free"
  ) +

  scale_fill_manual(
    values = c(
      "Surface/ambient" = "grey75",
      "Diatom lysate" = "#CC79A7",
      "75 m" = "#009E73",
      "200 m" = "#E69F00"
    ),
    name = NULL
  ) +

  scale_shape_manual(
    values = c(
      "homogeneous" = 21,
      "mixed" = 24
    ),
    labels = c(
      "homogeneous" = "Homogeneous (1 mm)",
      "mixed" = "Mixed (1 + 4 mm)"
    ),
    name = NULL
  ) +

  guides(
    fill = guide_legend(
      order = 1,
      title = NULL,
      override.aes = list(
        shape = 22,
        size = 4,
        color = "black",
        alpha = 1
      )
    ),

    shape = guide_legend(
      order = 2,
      title = NULL,
      override.aes = list(
        fill = "grey75",
        color = "grey25",
        size = 3,
        alpha = 1
      )
    )
  ) +

  labs(
    x = "Experiment",
    y = expression(
      "Single-column DOC removal (" *
        mu * "mol C L"^{-1} * ")"
    ),
    # caption = paste(
    #   "Small symbols show individual bioreactor columns;",
    #   "squares and error bars show mean ± SD among columns.",
    #   "The dotted line indicates the 1.5 µmol C L⁻¹",
    #   "analytical detection threshold."
    # )
  ) +

  theme_bw(base_size = 10.5) +

  theme(
    legend.position = "top",
    legend.box = "vertical",

    strip.background =
      element_rect(fill = "white"),

    strip.text =
      element_text(face = "bold"),

    axis.text.x = element_text(
      angle = 30,
      hjust = 1,
      vjust = 1
    ),

    panel.grid.minor =
      element_blank(),

    plot.caption = element_text(
      hjust = 0,
      size = 9
    ),

    plot.margin =
      margin(6, 10, 10, 6)
  )

fig3

ggsave(
  "figures/Figure_3_postconditioning.pdf",
  fig3,
  width = 11,
  height = 6.5
)
```

# Reporting tables

```{r reporting-tables}
figure2_reporting <- conditioning_summary %>%
  select(
    exp,
    date,
    days_conditioning,
    months_conditioning,

    influent_mean,
    influent_sd,
    influent_n_vials,

    mean_effluent_doc,
    sd_effluent_doc,
    effluent_n_vials_min,
    effluent_n_vials_max,

    n_bioreactors,
    mean_doc_removal,
    sd_doc_removal,
    n_exceeding_detection_limit
  ) %>%
  arrange(date)

figure3_reporting <- post_summary %>%
  select(
    exp,
    experiment_condition,
    date,
    site,
    condition,
    flow_rate_ml_min,
    flow_label,
    hours_elapsed,

    n_bioreactors,
    n_unique_bioreactor_labels,

    mean_doc_removal,
    sd_doc_removal,
    mean_pct_removal,
    sd_pct_removal,

    influent_n_vials,
    effluent_n_vials_min,
    effluent_n_vials_max
  ) %>%
  arrange(
    site,
    experiment_condition,
    condition,
    flow_rate_ml_min
  )

write_csv(
  figure2_reporting,
  "results/figure2_reporting_table.csv"
)

write_csv(
  figure3_reporting,
  "results/figure3_reporting_table.csv"
)
```


- DOC concentrations were calculated from replicate analytical vials; vial-level SD describes analytical variability.
- Individual bioreactor columns were treated as the experimental units for reactor-performance summaries.
- Figure error bars show the arithmetic mean ± SD among bioreactor columns.
- Figure `n` denotes the number of bioreactor columns, not the number of DOC vials.
- Influent and effluent vial counts are reported separately in the experiment metadata.
- BR15-4 included distinct 4 and 20 mL min-1 flow-rate treatments and these are analyzed separately.
- Four- and seven-hour assay durations reflect the time required to flush different reactor configurations several residence times before effluent collection.
- Conditioning results are interpreted as the onset of reproducible operational DOC removal; the study did not include recurring labile-carbon positive controls sufficient to attribute that transition exclusively to biofilm maturation.


# Supplemental Figure S2: BR15-1 time-course and bead configuration

This analysis retains individual bioreactor columns as the experimental units. DOC vial-level SD and SE are retained as analytical metadata but are not used to inverse-variance weight independent columns.

```{r figure-s2-data}
s2_base <- df %>%
  filter(
    exp == "BR15-1",
    type == "bioreactor",
    water == "UCSBintake",
    trt %in% c("flowthru", "DiatomExudate", "DiatomLysate"),
    hours_elapsed %in% c(0, 3, 4, 5, 6),
    stage %in% c(0, 1)
  ) %>%
  mutate(
    beads = tolower(trimws(beads)),
    beads = recode(
      beads,
      "homogenous" = "homogeneous",
      "mixed 1+4mm" = "mixed",
      .default = beads
    ),
    beads = factor(beads, levels = c("homogeneous", "mixed")),
    treatment = factor(
      trt_lab,
      levels = c("ambient", "diatom exudate", "diatom lysate")
    )
  )

s2_influent <- s2_base %>%
  filter(stage == 0, hours_elapsed == 0) %>%
  select(
    exp, date, water_lab, treatment,
    influent_mean = mean_doc,
    influent_sd = sd_doc,
    influent_se = se_doc,
    influent_n_vials = n_doc_vials
  ) %>%
  distinct()

s2_reactor <- s2_base %>%
  filter(stage == 1, hours_elapsed %in% c(3, 4, 5, 6)) %>%
  select(
    exp, date, water_lab, treatment,
    beads, bioreactor, flow_rate_ml_min, hours_elapsed,
    effluent_mean = mean_doc,
    effluent_sd = sd_doc,
    effluent_se = se_doc,
    effluent_n_vials = n_doc_vials
  ) %>%
  left_join(
    s2_influent,
    by = c("exp", "date", "water_lab", "treatment")
  ) %>%
 mutate(
  doc_removal = influent_mean - effluent_mean,
  delta_doc = effluent_mean - influent_mean,
  exceeds_detection_limit = doc_removal >= detection_limit
)

s2_summary <- s2_reactor %>%
  group_by(treatment, beads, hours_elapsed) %>%
  summarise(
    mean_effluent_doc = mean(effluent_mean),
    sd_effluent_doc = if_else(n() > 1, sd(effluent_mean), NA_real_),
    mean_doc_removal = mean(doc_removal),
    sd_doc_removal = if_else(n() > 1, sd(doc_removal), NA_real_),
    n_bioreactors = n(),
    influent_n_vials = first(influent_n_vials),
    effluent_n_vials_min = suppressWarnings(
      min(effluent_n_vials, na.rm = TRUE)
    ),
    effluent_n_vials_max = suppressWarnings(
      max(effluent_n_vials, na.rm = TRUE)
    ),
    .groups = "drop"
  ) %>%
  mutate(
    effluent_n_vials_min = if_else(
      is.infinite(effluent_n_vials_min), NA_real_, effluent_n_vials_min
    ),
    effluent_n_vials_max = if_else(
      is.infinite(effluent_n_vials_max), NA_real_, effluent_n_vials_max
    )
  )

write_csv(s2_reactor, "results/figureS2_reactor_level.csv")
write_csv(s2_summary, "results/figureS2_summary.csv")
```

```{r figure-s2, fig.width=8.5, fig.height=7}
bead_cols <- c(
  "homogeneous" = "#009E73",
  "mixed" = "#E69F00"
)

s2 <- ggplot(
  s2_reactor,
  aes(
    x = hours_elapsed,
    y = effluent_mean,
    group = interaction(treatment, beads, bioreactor)
  )
) +
  geom_hline(
    data = s2_influent,
    aes(yintercept = influent_mean),
    inherit.aes = FALSE,
    linetype = "dashed",
    linewidth = 0.45
  ) +
  geom_errorbar(
    data = s2_summary,
    aes(
      x = hours_elapsed,
      ymin = mean_effluent_doc - sd_effluent_doc,
      ymax = mean_effluent_doc + sd_effluent_doc,
      group = beads
    ),
    inherit.aes = FALSE,
    width = 0.10,
    position = position_dodge(width = 0.18),
    linewidth = 0.6,
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  geom_point(
    data = s2_summary,
    aes(
      x = hours_elapsed,
      y = mean_effluent_doc,
      group = beads
    ),
    inherit.aes = FALSE,
    shape = 22,
    fill = "white",
    color = "black",
    size = 3.3,
    stroke = 0.8,
    position = position_dodge(width = 0.18),
    show.legend = FALSE
  ) +
  geom_line(
    color = "grey65",
    alpha = 0.6,
    linewidth = 0.5,
    show.legend = FALSE
  ) +
  geom_point(
    aes(shape = beads),
    color = "grey30",
    size = 2.4,
    alpha = 0.8,
    position = position_jitter(width = 0.035, height = 0)
  ) +
  facet_wrap(
    ~ treatment,
    ncol = 1,
    scales = "free_y"
  ) +
  scale_shape_manual(
    values = c(
      "homogeneous" = 16,
      "mixed" = 17
    ),
    labels = c(
      "homogeneous" = "Homogeneous (1 mm)",
      "mixed" = "Mixed (1 + 4 mm)"
    ),
    name = NULL
  ) +
  guides(
    shape = guide_legend(
      title = NULL,
      override.aes = list(
        size = 2.8,
        color = "grey30",
        alpha = 1
      )
    )
  ) +
  scale_x_continuous(
    breaks = c(3, 4, 5, 6)
  ) +
  labs(
    x = "Elapsed assay time (h)",
    y = expression(
      "Effluent DOC (" * mu * "mol C L"^{-1} * ")"
    ),
    # caption = paste(
    #   "Dashed lines show the corresponding 0 h influent DOC.",
    #   "Small symbols and gray lines show individual bioreactor columns;",
    #   "open squares and error bars show mean ± SD among columns."
    # )
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position = "top",
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    plot.caption = element_text(hjust = 0, size = 9)
  )

s2

ggsave(
  "figures/Figure_S2_BR15_1_timecourse.pdf",
  s2,
  width = 8.5,
  height = 7
)
```

# Supplemental Figure S2 stability assessment

Because the same reactor columns were sampled repeatedly through time, stability comparisons should be based on within-column changes rather than z-tests of independently summarized means.

```{r figure-s2-stability}
stability_pairs <- tribble(
  ~hour1, ~hour2,
  3, 4,
  4, 5,
  4, 6
)

# One row per treatment × bead type × bioreactor,
# with sampling hours in separate columns.
s2_wide <- s2_reactor %>%
  select(
    treatment,
    beads,
    bioreactor,
    hours_elapsed,
    effluent_mean
  ) %>%
  distinct() %>%
  pivot_wider(
    names_from = hours_elapsed,
    values_from = effluent_mean,
    names_prefix = "hour_"
  )

# Add the three requested comparisons to every reactor row.
s2_pairwise <- s2_wide %>%
  cross_join(stability_pairs) %>%
  mutate(
    mean_hour1 = case_when(
      hour1 == 3 ~ hour_3,
      hour1 == 4 ~ hour_4,
      TRUE ~ NA_real_
    ),

    mean_hour2 = case_when(
      hour2 == 4 ~ hour_4,
      hour2 == 5 ~ hour_5,
      hour2 == 6 ~ hour_6,
      TRUE ~ NA_real_
    ),

    change_doc = mean_hour2 - mean_hour1
  ) %>%
  filter(
    is.finite(mean_hour1),
    is.finite(mean_hour2),
    is.finite(change_doc)
  )

s2_pairwise_summary <- s2_pairwise %>%
  group_by(treatment, beads, hour1, hour2) %>%
  summarise(
    n_bioreactors = n(),
    mean_change_doc = mean(change_doc),
    sd_change_doc = if_else(n() > 1, sd(change_doc), NA_real_),
    median_change_doc = median(change_doc),
    max_abs_change_doc = max(abs(change_doc)),
   all_changes_below_analytical_threshold =
  all(abs(change_doc) < detection_limit),
    .groups = "drop"
  )

# Optional paired t-tests. Treat as exploratory because n is small.
s2_pairwise_tests <- s2_pairwise %>%
  group_by(treatment, beads, hour1, hour2) %>%
  group_modify(~{
    if (nrow(.x) >= 2 && sd(.x$change_doc) > 0) {
      tt <- t.test(.x$change_doc, mu = 0)
      tibble(
        n_bioreactors = nrow(.x),
        mean_change_doc = mean(.x$change_doc),
        sd_change_doc = sd(.x$change_doc),
        t_value = unname(tt$statistic),
        df = unname(tt$parameter),
        p_value = tt$p.value
      )
    } else {
      tibble(
        n_bioreactors = nrow(.x),
        mean_change_doc = mean(.x$change_doc),
        sd_change_doc = if_else(nrow(.x) > 1, sd(.x$change_doc), NA_real_),
        t_value = NA_real_,
        df = NA_real_,
        p_value = NA_real_
      )
    }
  }) %>%
  ungroup() %>%
  mutate(
  mean_change_exceeds_analytical_threshold =
    abs(mean_change_doc) >= detection_limit
)

write_csv(
  s2_pairwise_summary,
  "results/figureS2_stability_summary.csv"
)

write_csv(
  s2_pairwise_tests,
  "results/figureS2_stability_tests.csv"
)
```

# Supplemental Figure S3: bead-pack comparison

This analysis compares single-column DOC removal between homogeneous and mixed bead packs. Each point is an individual bioreactor column, not an inverse-variance-weighted run mean. Flow rate is included explicitly because BR15-4 contained distinct 4 and 20 mL min-1 conditions.

```{r figure-s3-data}
exp_15 <- paste0("BR15-", 1:11)

s3_base <- df %>%
  filter(
    exp %in% exp_15,
    type == "bioreactor",
    stage %in% c(0, 1),
    hours_elapsed %in% c(0, 4)
  ) %>%
  mutate(
    beads = tolower(trimws(beads)),
    beads = recode(
      beads,
      "homogenous" = "homogeneous",
      "mixed 1+4mm" = "mixed",
      .default = beads
    ),
    beads = factor(beads, levels = c("homogeneous", "mixed"))
  )

s3_influent <- s3_base %>%
  filter(stage == 0, hours_elapsed == 0) %>%
  select(
    exp, date, water_lab, trt_lab,
    influent_doc = mean_doc,
    influent_sd = sd_doc,
    influent_se = se_doc,
    influent_n_vials = n_doc_vials
  ) %>%
  distinct()

s3_reactor <- s3_base %>%
  filter(stage == 1, hours_elapsed == 4) %>%
  select(
    exp, date, water_lab, trt_lab,
    beads, bioreactor, flow_rate_ml_min,
    effluent_doc = mean_doc,
    effluent_sd = sd_doc,
    effluent_se = se_doc,
    effluent_n_vials = n_doc_vials
  ) %>%
  left_join(
    s3_influent,
    by = c("exp", "date", "water_lab", "trt_lab")
  ) %>%
  filter(is.finite(influent_doc), is.finite(effluent_doc)) %>%
  mutate(
    removal_doc = influent_doc - effluent_doc,
    pct_removal = 100 * removal_doc / influent_doc,
    flow_rate_factor = factor(flow_rate_ml_min)
  )

s3_summary <- s3_reactor %>%
  group_by(
    exp,
    date,
    water_lab,
    trt_lab,
    beads,
    flow_rate_ml_min
  ) %>%
  summarise(
    influent_doc = first(influent_doc),
    influent_sd = first(influent_sd),
    influent_n_vials = first(influent_n_vials),

    mean_effluent_doc = mean(effluent_doc, na.rm = TRUE),
    sd_effluent_doc = if_else(
      sum(is.finite(effluent_doc)) > 1,
      sd(effluent_doc, na.rm = TRUE),
      NA_real_
    ),

    mean_removal_doc = mean(removal_doc, na.rm = TRUE),
    sd_removal_doc = if_else(
      sum(is.finite(removal_doc)) > 1,
      sd(removal_doc, na.rm = TRUE),
      NA_real_
    ),

    n_bioreactors = n(),
    .groups = "drop"
  )

write_csv(
  s3_summary,
  "results/figureS3_summary.csv"
)

write_csv(s3_reactor, "results/figureS3_reactor_level.csv")
```

```{r figure-s3-models}

s3_model_audit <- s3_reactor %>%
  summarise(
    n_total = n(),
    n_with_flow_rate = sum(is.finite(flow_rate_ml_min)),
    n_missing_flow_rate = sum(is.na(flow_rate_ml_min))
  )

print(s3_model_audit)

# Primary model: individual reactor columns are observations.
# Account for source/treatment heterogeneity and flow rate.
mod_s3_full <- lm(
  removal_doc ~ beads + influent_doc +
    water_lab + trt_lab + flow_rate_ml_min,
  data = s3_reactor
)

mod_s3_interaction <- lm(
  removal_doc ~ beads * influent_doc +
    water_lab + trt_lab + flow_rate_ml_min,
  data = s3_reactor
)

table_s3_lm_coefficients <- broom::tidy(
  mod_s3_full,
  conf.int = TRUE
)

table_s3_anova <- broom::glance(mod_s3_full)

# Bead-pack comparison adjusted for covariates.
# Type-II tests avoid dependence on predictor entry order.
if (requireNamespace("car", quietly = TRUE)) {
  table_s3_type2 <- car::Anova(mod_s3_full, type = 2) %>%
    as.data.frame() %>%
    rownames_to_column("term")
} else {
  table_s3_type2 <- tibble(
    note = "Install the car package to calculate Type-II tests."
  )
}

# Sensitivity analysis excluding the 20 mL min-1 BR15-4 treatment.
mod_s3_4ml <- lm(
  removal_doc ~ beads * influent_doc + water_lab + trt_lab,
  data = s3_reactor %>%
    filter(is.na(flow_rate_ml_min) | abs(flow_rate_ml_min - 4) < 0.1)
)

table_s3_4ml_coefficients <- broom::tidy(
  mod_s3_4ml,
  conf.int = TRUE
)

write_csv(
  broom::tidy(mod_s3_full, conf.int = TRUE),
  "results/figureS3_primary_model_coefficients.csv"
)

write_csv(
  broom::tidy(mod_s3_interaction, conf.int = TRUE),
  "results/figureS3_interaction_model_coefficients.csv"
)

write_csv(
  table_s3_anova,
  "results/figureS3_full_model_glance.csv"
)

write_csv(
  table_s3_type2,
  "results/figureS3_full_model_type2.csv"
)

write_csv(
  table_s3_4ml_coefficients,
  "results/figureS3_4ml_sensitivity_coefficients.csv"
)
```

```{r figure-s3, fig.width=8.2, fig.height=5.6}
s3_plot <- ggplot(
  s3_reactor,
  aes(x = influent_doc, y = removal_doc)
) +
  geom_hline(
    yintercept = 0,
    linewidth = 0.4,
    color = "grey60"
  ) +
  geom_hline(
    yintercept = detection_limit,
    linetype = "dotted",
    linewidth = 0.4
  ) +
  geom_point(
    aes(fill = beads, shape = flow_rate_factor),
    color = "black",
    stroke = 0.5,
    size = 3,
    alpha = 0.85
  ) +
  geom_smooth(
    aes(color = beads, fill = beads),
    method = "lm",
    se = TRUE,
    linewidth = 0.9,
    alpha = 0.15
  ) +
  scale_fill_manual(
    values = bead_cols,
    labels = c(
      "homogeneous" = "Homogeneous (1 mm)",
      "mixed" = "Mixed (1 + 4 mm)"
    ),
    name = "Bead configuration"
  ) +
  scale_color_manual(
    values = bead_cols,
    labels = c(
      "homogeneous" = "Homogeneous (1 mm)",
      "mixed" = "Mixed (1 + 4 mm)"
    ),
    name = "Bead configuration"
  ) +
  labs(
    x = expression(
      "Initial influent DOC (" * mu * "mol C L"^{-1} * ")"
    ),
    y = expression(
      "Single-column DOC removal (" * mu * "mol C L"^{-1} * ")"
    ),
    shape = expression("Flow rate (mL min"^{-1} * ")"),
    caption = paste(
      "Points are individual bioreactor columns.",
      "Lines are unadjusted bead-specific OLS fits with 95% confidence intervals;",
      "the accompanying model adjusts for water source, treatment, and flow rate."
    )
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position = "top",
    panel.grid.minor = element_blank()
  ) + 
  guides(
  color = "none",
  fill = guide_legend(
    order = 1,
    title = "Bead configuration",
    override.aes = list(shape = 21, size = 3.5)
  ),
  shape = guide_legend(
    order = 2,
    title = expression("Flow rate (mL min"^{-1} * ")")
  )
)

s3_plot

ggsave(
  "figures/Figure_S3_bead_comparison.pdf",
  s3_plot,
  width = 8.2,
  height = 5.6
)
```

